---
title: "aab"
output: html_document
date: "2025-04-02"
---

This script is dedicated to the analysis of autoantibody (AAB) data. Specifically, it examines samples from 33 patients diagnosed with Primary Sclerosing Cholangitis (PSC). For the purposes of this study, patients were stratified into five clinical groups based on disease characteristics: Early-stage PSC, Advanced-stage PSC, PSC without associated Inflammatory Bowel Disease (no IBD), PSC with Cholangiocarcinoma (CCA), and Progressive PSC.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Importation of all necessary packages required for the analysis. These libraries provide essential functions for data manipulation, visualization, and statistical evaluation throughout the study.

```{r}
# Data manipulation and visualization packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(naniar)
library(gridExtra)
library(viridis)
library(pbapply)
library(FactoMineR)
library(factoextra)
library(pheatmap)
library(rstatix)
library(ggpubr)

# Packages for enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)

```

# 1. Explonatory Analysis

This section performs a broad initial inspection of the dataset to understand overall distributions, missingness patterns, and variability before moving on to more targeted tests.

```{r}
#loading the data
aab <- read.csv("aab_clean.csv")
```

```{r}
# Reshape data to long format for easier analysis
aab_long <- aab %>%
  pivot_longer(-group, names_to = "protein", values_to = "expression")
```

## Distribution by Group

We visualise expression distributions within each group to spot shifts, skewness or outliers that could signal biological differences or data quality issues.

```{r}
# Histogram: distribution of expression values per group
ggplot(aab_long, aes(x = expression, fill = group)) +
  geom_histogram(alpha = 0.7, bins = 50) +
  facet_wrap(~group) +
  scale_fill_viridis(discrete = TRUE) +
  xlim(-20, 20) +
  theme_minimal() +
  labs(title = "Expression Distribution by Group",
       x = "Expression", y = "Frequency")

# Density plots: overlapping expression density per group
ggplot(aab_long, aes(x = expression, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis(discrete = TRUE) +
  xlim(-20, 20) +
  theme_minimal() +
  labs(title = "Expression Density by Group",
       x = "Expression", y = "Density")

# Boxplot: comparative expression across groups
ggplot(aab_long, aes(x = group, y = expression, fill = group)) +
  geom_boxplot() +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Expression Boxplot by Group",
       x = "Group", y = "Expression")


```

## Missing Data Exploration

Understanding where data are missing and how much is missing helps decide whether to impute, filter, or otherwise handle incomplete observations.

```{r}
# Calculate missing data percentages by group
missing_by_group <- aab_long %>%
  group_by(group) %>%
  summarise(
    n_total = n(),
    n_missing = sum(is.na(expression)),
    pct_missing = 100 * n_missing / n_total
  )

# Barplot of missing percentages with count labels
ggplot(missing_by_group, aes(x = group, y = pct_missing, fill = group)) +
  geom_col() +
  geom_text(aes(label = n_missing), position = position_stack(vjust = 0.5), color = "white") +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Percentage of Missing Values by Group",
       x = "Group", y = "% Missing")

# Heatmap of missing values: use a random sample of proteins for clarity
set.seed(123)
sample_proteins <- sample(unique(aab_long$protein), 100)

data_subset <- aab_long %>%
  filter(protein %in% sample_proteins) %>%
  pivot_wider(names_from = protein, values_from = expression)

vis_miss(data_subset, sort_miss = TRUE) +
  labs(title = "Missing Values Visualization (Sample of 100 Proteins)")


```

## Descriptive Statistics

Key summary metrics (mean, median, variance, etc.) by group provide a quick quantitative check of central tendency and spread, supporting later hypothesis tests.

```{r}
# Compute descriptive statistics per group
stats_by_group <- aab_long %>%
  group_by(group) %>%
  summarise(
    n = sum(!is.na(expression)),
    mean = mean(expression, na.rm = TRUE),
    median = median(expression, na.rm = TRUE),
    sd = sd(expression, na.rm = TRUE),
    cv = 100 * sd / abs(mean),  # Coefficient of Variation (%)
    min = min(expression, na.rm = TRUE),
    max = max(expression, na.rm = TRUE),
    q25 = quantile(expression, 0.25, na.rm = TRUE),
    q75 = quantile(expression, 0.75, na.rm = TRUE)
  )

# Display summary stats using kable
knitr::kable(stats_by_group, caption = "Descriptive Statistics by Group", digits = 3)

# Bar plots for key statistics across groups
p1 <- ggplot(stats_by_group, aes(x = group, y = mean, fill = group)) +
  geom_col() +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Mean Expression by Group", y = "Mean Expression") +
  theme(legend.position = "none")

p2 <- ggplot(stats_by_group, aes(x = group, y = sd, fill = group)) +
  geom_col() +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Standard Deviation by Group", y = "SD") +
  theme(legend.position = "none")

p3 <- ggplot(stats_by_group, aes(x = group, y = cv, fill = group)) +
  geom_col() +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Coefficient of Variation by Group", y = "CV (%)") +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, ncol = 3)
```

## Counting Proteins with Zero Missing Values

Calculating how many proteins are fully observed informs the feasibility of analyses that cannot accommodate missing data.

```{r}
# Total unique proteins
n_total_proteins <- length(unique(aab_long$protein))

# Count proteins with no missing values
df_complete <- aab_long %>%
  group_by(protein) %>%
  filter(!any(is.na(expression))) %>%
  ungroup()

n_complete_proteins <- length(unique(df_complete$protein))
percent_retained <- n_complete_proteins / n_total_proteins * 100

cat(sprintf("Initial proteins: %d\nProteins with no NA: %d (%.1f%%)\n", 
            n_total_proteins, n_complete_proteins, percent_retained))

```

## Scatter Plots Between Groups

Pairwise scatter plots reveal concordance or systematic shifts between groups and flag proteins that deviate beyond a set threshold.

```{r}

# Transform data to wide format and remove rows with any missing values
aab_wide <- aab_long %>%
  pivot_wider(id_cols = protein, names_from = group, values_from = expression) %>%
  drop_na()

# Extract group names (exclude 'protein' column)
group_names <- names(aab_wide)[-1]

# Create all pairwise combinations (2 at a time) among groups
combs <- combn(group_names, 2, simplify = FALSE)

# Initialize list for storing plots
plot_list <- list()

# Generate scatter plots for each pair of groups with decision boundaries
for(pair in combs){
  
  # Flag points outside the region defined by +/- 10 difference between groups
  df_temp <- aab_wide %>% 
    mutate(status = ifelse((.data[[pair[2]]] > .data[[pair[1]]] + 10) | 
                           (.data[[pair[2]]] < .data[[pair[1]]] - 10),
                           "outside", "inside"))
  
  # Create scatter plot with diagonal lines at intercepts +/- 10
  p <- ggplot(df_temp, aes_string(x = pair[1], y = pair[2], color = "status")) +
    geom_point(size = 1.5) +
    geom_abline(intercept = 10, slope = 1, linewidth = 0.1, linetype = "solid", color = "black") +
    geom_abline(intercept = -10, slope = 1, linewidth = 0.1, linetype = "solid", color = "black") +
    scale_color_manual(values = c("inside" = "steelblue", "outside" = "#f9e13a")) +
    scale_x_continuous(limits = c(-60, 60), breaks = seq(-60, 60, by = 20)) +
    scale_y_continuous(limits = c(-60, 60), breaks = seq(-60, 60, by = 20)) +
    theme_bw() +
    theme(panel.background = element_rect(fill = "#e5ecf6", color = NA),
          plot.background = element_rect(fill = "white", color = NA),
          panel.grid.major = element_line(color = "white"),
          panel.grid.minor = element_line(color = "white")) +
    labs(x = pair[1], y = pair[2],
         title = paste("Scatter Plot:", pair[1], "vs", pair[2]))
  
  # Save each plot to a PNG file
  plot_name <- paste(pair, collapse = "_")
  plot_list[[plot_name]] <- p
  ggsave(
    filename = paste0("../results/scatter_", pair[1], "_", pair[2], ".png"),
    plot = p,
    width = 5, height = 4, dpi = 300,
    bg = "white"
  )
}

# Arrange scatter plots in a mosaic (2 columns)
grid.arrange(grobs = plot_list, ncol = 2)

```

# 2. In-depth Analyses

Advanced multivariate and specificity analysis to extract biologically meaningful patterns.

### Z-score Normalization

Puts every protein on the same scale so profiles can be compared directly.

```{r}

# Remove the 'protein' column and perform row-wise z-score normalization
aab_wide_data <- aab_wide[, -1]
zscore_norm <- t(apply(aab_wide_data, 1, function(x) scale(x)))

# Restore column names and add the protein identifier
colnames(zscore_norm) <- colnames(aab_wide_data)
aab_wide_zscore <- data.frame(protein = aab_wide$protein, zscore_norm)


```

```{r}
# Convert the z-score normalized wide-format data to long format
aab_long_zscore <- aab_wide_zscore %>%
  pivot_longer(
    cols = starts_with("Group"),
    names_to = "group",
    values_to = "expression"
  )

# Histogram: z-score expression distribution by group
ggplot(aab_long_zscore, aes(x = expression, fill = group)) +
  geom_histogram(alpha = 0.7, bins = 50) +
  facet_wrap(~ group) +
  scale_fill_viridis(discrete = TRUE) +
  xlim(-20, 20) +
  theme_minimal() +
  labs(title = "Z-score Expression Distribution by Group",
       x = "Z-score", y = "Frequency")

# Density plots: overlapping densities
ggplot(aab_long_zscore, aes(x = expression, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_viridis(discrete = TRUE) +
  xlim(-20, 20) +
  theme_minimal() +
  labs(title = "Z-score Expression Density by Group",
       x = "Z-score", y = "Density")

# Boxplot: comparative z-score per group
ggplot(aab_long_zscore, aes(x = group, y = expression, fill = group)) +
  geom_boxplot() +
  scale_fill_viridis(discrete = TRUE) +
  theme_minimal() +
  labs(title = "Z-score Expression Boxplot by Group",
       x = "Group", y = "Z-score")


```

### PCA

Compresses the data to two axes to see how well clinical groups separate.

```{r}
# Transpose the normalized data so that each group is an observation
data_matrix <- aab_wide_zscore[, -1] %>% as.data.frame() %>% t() %>% as.data.frame()

# Rename rows using the original group names and assign protein IDs to columns
group_names <- colnames(aab_wide_zscore)[-1]
group_names <- recode(group_names,
                      "Group1" = "Advanced",
                      "Group2" = "Early",
                      "Group3" = "No IBD",
                      "Group4" = "CCA",
                      "Group5" = "Progressor")
rownames(data_matrix) <- group_names
colnames(data_matrix) <- aab_wide_zscore$protein

# Perform PCA (data are already normalized per protein)
pca_result <- prcomp(data_matrix, center = TRUE, scale. = FALSE)
explained_var <- (pca_result$sdev^2) / sum(pca_result$sdev^2) * 100
pc1_perc <- round(explained_var[1], 1)
pc2_perc <- round(explained_var[2], 1)

# Assemble PCA scores and plot results
pca_df <- as.data.frame(pca_result$x)
pca_df$group <- rownames(pca_df)

ggplot(pca_df, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 4) +
  geom_text(aes(label = group), vjust = -0.8, show.legend = FALSE) +
  labs(title = "PCA of Groups", x = paste0("PC1 (", pc1_perc, "%)"),
       y = paste0("PC2 (", pc2_perc, "%)")) +
  theme_minimal() +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))
```

### Hierarchical Clustering

Checks if samples cluster by group without any supervision.

```{r}
# Remove the protein column and rename group columns
aab_data <- aab_wide_zscore[, -1]
colnames(aab_data) <- recode(colnames(aab_data),
                             "Group1" = "Advanced",
                             "Group2" = "Early",
                             "Group3" = "No IBD",
                             "Group4" = "CCA",
                             "Group5" = "Progressor")

# Transpose the data and compute Euclidean distance between groups
aab_t <- t(aab_data)
distance_matrix <- dist(aab_t, method = "euclidean")

# Perform hierarchical clustering using Ward's method and plot the dendrogram
hc <- hclust(distance_matrix, method = "ward.D2")
plot(hc, main = "Hierarchical Clustering of Groups", xlab = "Groups", sub = "", cex = 0.9)

```

### Identify Best Proteins per Group

Scores proteins for high signal in one group and low signal elsewhere.

1)  Positive signal from the target group (signal_target)

-   If the value of the target group \> 0: keep it as is.
-   Otherwise: set it to 0 (considered as "no signal").

2)  Highest positive signal among the 4 other groups (max_others)

-   Check the 4 remaining groups; keep only values \> 0.
-   Take the highest one.
-   If all values are ≤ 0, then max_others = 0.

3)  Average of positive signals in the 4 other groups (avg_others)

-   Calculate the average of the positive values (0 if none).

4)  Distance to the closest competitor (diff_max)

-   diff_max = signal_target − max_others
-   Measures how much the protein exceeds the highest positive value in the other groups.

5)  Specificity factor (τ)

-   If signal_target = 0: τ = 0 (not specific).
-   Otherwise: τ = 1 − (avg_others / signal_target), which lies between 0 and 1.
-   The higher the average of the other groups, the lower τ becomes.
-   The idea is to penalize cases where the target group is very high but another group is also high, while favoring cases where only the target group is high.

6)  Final score

-   score = max(diff_max, 0) × τ
-   Combines signal strength (diff_max) and specificity (τ).
-   A high score means: strong signal in the target group and almost no positive signals in the others.

7)  Ranking

-   Sort all proteins by decreasing score.
-   Keep the top 300 for each group.

```{r}
group_cols   <- setdiff(names(aab_wide), "protein")      # all group columns
result_list  <- vector("list", length(group_cols))       # pre-allocate output
names(result_list) <- group_cols

for (grp in group_cols) {
  other_cols <- setdiff(group_cols, grp)                # non-focal groups
  
  result_list[[grp]] <- aab_wide %>% 
    rowwise() %>% 
    mutate(
      # 1. Positive signal of the focal group
      pos_focal      = if_else(.data[[grp]] > 0, .data[[grp]], 0),
      
      # 2. Strongest positive signal in the remaining groups
      max_pos_other  = max(if_else(c_across(all_of(other_cols)) > 0,
                                   c_across(all_of(other_cols)), 0),
                           na.rm = TRUE),
      
      # 3. Mean positive signal in the remaining groups
      mean_pos_other = mean(if_else(c_across(all_of(other_cols)) > 0,
                                    c_across(all_of(other_cols)), 0),
                            na.rm = TRUE),
      
      # 4. Distance to the closest competitor
      diff_max = pos_focal - max_pos_other,
      
      # 5. Specificity factor τ
      tau      = if_else(pos_focal == 0,
                         0,
                         1 - mean_pos_other / pos_focal),
      
      # 6. Final score: intensity × specificity
      score    = pmax(diff_max, 0) * tau
    ) %>% 
    ungroup() %>% 
    filter(score > 0) %>%           # keep proteins where the focal group dominates
    arrange(desc(score)) %>%        # rank by decreasing score
    slice_head(n = 300)             # retain top 300 proteins for this group
}

# Example: six most specific proteins for “Group1”
head(result_list[["Group1"]], 6)
```

### Heatmap of Top Proteins

Visual snapshot of expression for the most group-specific proteins.

```{r heatmap, fig.width=30, fig.height=25, out.width='100%'}

# Generate a heatmap for the top proteins per group
for (grp in names(result_list)) {
  df_top <- result_list[[grp]]
  # Create matrix from expression columns, set row names as protein IDs
  mat <- as.matrix(df_top[, group_cols])  # Corrected variable name
  
  rownames(mat) <- df_top$protein
  
  pheatmap(mat,
           main = paste("Heatmap of Top 100 Specific Proteins -", grp),
           scale = "row",                       
           clustering_distance_rows = "euclidean", 
           clustering_distance_cols = "euclidean",
           clustering_method = "complete")
}
```

### Enrichment Analysis Preparation

Maps protein IDs to clean gene lists before GO testing.

```{r}
# 1. Read and prepare metadata
metadata <- read.csv("data_aab_metadata.csv", 
                     sep = ";", 
                     stringsAsFactors = FALSE)

# Clean the metadata:
metadata$Gene <- trimws(gsub('"', '', metadata$Gene))
metadata$PrEST <- trimws(metadata$PrEST)

# 2. For each dataframe in result_list, extract the PrEST (portion before "_"),
# join with metadata, and split multiple genes into separate rows
liste_enrch <- list()

for (grp in names(result_list)) {
  
  # Retrieve the current group dataframe from result_list and extract PrEST
  df <- result_list[[grp]] %>%
    mutate(PrEST = sub("_.*", "", protein)) %>%
    inner_join(metadata, by = "PrEST") %>%
    separate_rows(Gene, sep = ";") %>%
    dplyr::select(PrEST, Gene)
  
  # Store the result in the list with its group name
  liste_enrch[[grp]] <- df
}


```

### Enrichment Analysis Execution

Finds Biological-Process terms over-represented in each protein set.

```{r}
if (!dir.exists("ORA_aab")) {
  dir.create("ORA_aab")
}

for (grp in names(liste_enrch)) {
  cat("[1] \"", grp, "\"\n", sep = "")
  geneList <- unique(liste_enrch[[grp]]$Gene)
  cat("Number of genes for", grp, ":", length(geneList), "\n")
  
  if(length(geneList) == 0) {
    cat("No genes available for group:", grp, "\n")
    next
  }
  
  enrich_result <- tryCatch({
    enrichGO(
      gene         = geneList,
      OrgDb        = org.Hs.eg.db,
      keyType      = "SYMBOL",
      ont          = "BP",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2
    )
  }, error = function(e) {
    message("enrichGO error in group ", grp, ": ", e$message)
    return(NULL)
  })
  
  if(is.null(enrich_result)) {
    message("No enrichment result for group: ", grp)
    next
  }
  
  res_df <- as.data.frame(enrich_result)
  
  if(nrow(res_df) == 0) {
    cat("No enrichment found for group:", grp, "\n")
  } else {
    cat("Enrichment found for group:", grp, "\n")
    p <- dotplot(enrich_result, showCategory = 20) +
         ggtitle(paste("GO BP Enrichment for", grp))
    ggsave(filename = paste0("ORA_aab/", grp, ".png"), plot = p, width = 8, height = 6)
    cat("Plot saved for group:", grp, "\n")
  }
  
  cat("Finished processing group:", grp, "\n\n")
}

```

### Bonus: Top‑Signal Proteins Without Specificity Penalty

A complementary ranking based purely on strongest signals (no specificity penalty) to capture universally high responders

```{r}
# ╭─────────────────────────────────────────────────────────╮
# │  1. Top 300 raw (strongest signals, no penalties)        │
# ╰─────────────────────────────────────────────────────────╯
result_list_2 <- list()

for (grp in group_cols) {                         
  result_list_2[[grp]] <- aab_wide %>%            # aab_wide_zscore
    arrange(desc(.data[[grp]])) %>%               # sort by group value
    slice_head(n = 300)                           # keep top 300 rows
}

# ╭─────────────────────────────────────────────────────────╮
# │  2. Preparation of gene sets                             │
# ╰─────────────────────────────────────────────────────────╯
liste_enrch2 <- list()

for (grp in names(result_list_2)) {
  liste_enrch2[[grp]] <- result_list_2[[grp]] %>%
    mutate(PrEST = sub("_.*", "", protein)) %>%          # extract PrEST ID
    inner_join(metadata, by = "PrEST") %>%               # add gene annotations
    separate_rows(Gene, sep = ";") %>%                   # one gene per row
    dplyr::select(PrEST, Gene)
}

# ╭─────────────────────────────────────────────────────────╮
# │  3. GO enrichment and saving of plots                   │
# ╰─────────────────────────────────────────────────────────╯
dir.create("ORA_aab_top300", showWarnings = FALSE)

for (grp in names(liste_enrch2)) {
  cat("[Top300] ", grp, "\n", sep = "")
  geneList <- unique(liste_enrch2[[grp]]$Gene)
  cat("   Number of genes: ", length(geneList), "\n")
  if (length(geneList) == 0) next

  enr <- tryCatch(
    enrichGO(
      gene         = geneList,
      OrgDb        = org.Hs.eg.db,
      keyType      = "SYMBOL",
      ont          = "BP",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.20
    ),
    error = function(e) { message("   enrichGO error: ", e$message); NULL }
  )
  if (is.null(enr) || nrow(as.data.frame(enr)) == 0) {
    cat("   No enrichment found\n\n")
    next
  }

  p <- dotplot(enr, showCategory = 20) +
       ggtitle(paste("GO BP – Top 300 –", grp))
  ggsave(paste0("ORA_aab_top300/", grp, "_top300.png"), p,
         width = 8, height = 6)
  cat("   Plot saved\n\n")
}
```

# 3. Link AAB with proteomic data

Validates antibody signals against independent mass-spectrometry data.

## 1) Correlation of 300 Group‑specific Proteins to the same Group Means

Ranks proteins by cosine similarity between AAB signatures and proteomic group means.

```{r}

library(dplyr)
# Combine all data frames in 'result_list' into a single data frame
# 'df_spe_complete' will have about 1,500 rows
df_spe_complete <- bind_rows(result_list)

# Create a final data frame by selecting specific columns
# Here, we remove ('-') the columns: pos_cible, max_pos_oth, mean_pos_oth, diff_max, tau, and score
df_spe_final <- df_spe_complete %>% 
 dplyr::select( -pos_focal,
                -max_pos_other,
                -mean_pos_other,
                -diff_max,
                -tau,
                -score)
```

```{r}
# Load the preprocessed protein dataset
protein <- read.csv("protein_preprocessed.csv")

# 1) Convert to data frame and keep only the first 33 rows
df2_proteo <- as.data.frame(protein)[1:33, ]

# 2) Create a grouping vector based on row indices
grp <- cut(seq_len(nrow(df2_proteo)),     
           breaks  = c(0, 7, 13, 21, 27, 33),  # Define the breakpoints for groups
           labels  = paste0("group", 1:5))      # Assign group labels

# 3) Aggregate the data by group
# Exclude the first column (patient_id) and calculate the mean for each group
df_proteo_merged <- aggregate(df2_proteo[ , -1], 
                       by   = list(group = grp),
                       FUN  = mean,
                       na.rm = TRUE)
```

This procedure evaluates the shape similarity (cosine similarity) between two datasets, df_spe_final and df_proteo_merged.

```{r}

## CORRELATION MATRIX (df_spe_final vs df_proteo_merged)

suppressPackageStartupMessages(library(dplyr))   # Load dplyr last to avoid conflicts

## 1) Extract matrices ---------------------------------------------------------
grp_cols  <- grep("^Group\\d+$", names(df_spe_final), value = TRUE)
mat_final <- as.matrix(df_spe_final[, grp_cols])
rownames(mat_final) <- df_spe_final$protein

group_names <- tolower(df_proteo_merged[[1]])           # "group1"…"group5"
mat_groups  <- t(as.matrix(df_proteo_merged[, -1]))
rownames(mat_groups) <- colnames(df_proteo_merged)[-1]
colnames(mat_groups) <- group_names

## 2) Harmonize matrices -------------------------------------------------------
col_order           <- tolower(grp_cols)       # "group1"…"group5"
colnames(mat_final) <- col_order
mat_groups          <- mat_groups[, col_order, drop = FALSE]
```

1)  Matrix Extraction

-   Extract group columns (Group1, Group2, etc.) from df_spe_final to create mat_final.

-   Transpose df_proteo_merged to create mat_groups, aligning groups as columns.

2)  Harmonize row and column names to lowercase.

-   Column Harmonization

-   Keep only common columns between mat_final and mat_groups (typically group1 to group5).

3)  Standard Deviation Filtering (σ \> min_sd)

-   Remove proteins with a standard deviation lower than min_sd = 0.5.

-   This removes proteins with low variability.

4)  Option: Using Differences (slopes)

-   If use_differences = TRUE, replace the group values by successive differences.

5)  Matrix Normalization

-   If center_only = TRUE: center each protein by subtracting its mean.

-   Else: apply Z-score normalization (center and scale by standard deviation).

6)  Shape Similarity Calculation

-   Compute the cosine similarity (normalized dot product) between each pair of proteins.

-   Generate the similarity matrix (sim_mat) with values between -1 and 1.

7)  Amplitude Bonus Calculation

-   Calculate an amplitude bonus based on comparing standard deviations between proteins.

-   Modulate its impact using the beta parameter.

8)  Combined Scoring

-   Calculate the final score:

-   score = similarity × (amplitude_factor)\^beta

-   A higher score indicates stronger similarity.

9)  Final Table Creation

-   Transform the similarity matrix into a tidy table (res_final).

-   Filter out low scores based on score_cutoff.

-   Sort by decreasing score.

10) Export and Visualization

-   Save res_final as a TSV file.

Generate individual line plots for each protein pair to visualize group profiles.

```{r}

## 0. PARAMETERS TO ADJUST

min_sd         <- 0.50   # Minimum standard deviation to keep a protein
score_cutoff   <- 0.75   # Minimum combined score threshold
beta           <- 0.1     # Weight for the amplitude bonus
use_differences <- TRUE # TRUE = use differences (slopes) instead of raw values
center_only     <- TRUE # TRUE = center values without scaling to unit variance


## 1. LOAD REQUIRED PACKAGES

suppressPackageStartupMessages(library(data.table))


## 2. ALIGN COLUMNS

# Keep only common group columns between the two matrices
common_groups <- intersect(colnames(mat_final), colnames(mat_groups))
mat_final_raw  <- as.matrix(mat_final[, common_groups])
mat_groups_raw <- as.matrix(mat_groups[, common_groups])


## 3. FILTER BASED ON STANDARD DEVIATION

# Remove proteins with low variability
keep_final  <- apply(mat_final_raw,  1, sd, na.rm = TRUE) > min_sd
keep_groups <- apply(mat_groups_raw, 1, sd, na.rm = TRUE) > min_sd

mat_final  <- mat_final_raw[keep_final , , drop = FALSE]
mat_groups <- mat_groups_raw[keep_groups, , drop = FALSE]


## 4. OPTIONAL: COMPUTE DIFFERENCES (SLOPES)

if (use_differences) {
  mat_final  <- t(apply(mat_final,  1, diff))  # n x (g-1)
  mat_groups <- t(apply(mat_groups, 1, diff))
}


## 5. NORMALIZATION: CENTERING OR SCALING

if (center_only) {
  center_rows <- function(m) t(apply(m, 1, function(v) v - mean(v, na.rm = TRUE)))
  X <- center_rows(mat_final)
  Y <- center_rows(mat_groups)
} else {
  scale_rows <- function(m) t(apply(m, 1, scale))
  X <- scale_rows(mat_final)
  Y <- scale_rows(mat_groups)
}


## 6. SHAPE SIMILARITY (COSINE SIMILARITY)

num   <- X %*% t(Y)
denom <- outer(sqrt(rowSums(X^2)), sqrt(rowSums(Y^2)))
sim_mat <- num / denom  # n_final × n_groups matrix


## 7. AMPLITUDE BONUS

sd_final  <- apply(mat_final,  1, sd)
sd_groups <- apply(mat_groups, 1, sd)

amp_sim    <- function(s1, s2) pmin(s1, s2) / pmax(s1, s2)
amp_factor <- outer(sd_final, sd_groups, amp_sim)  # Values between 0 and 1


## 8. COMBINED SCORE (SHAPE × AMPLITUDE^β)

score_mat <- sim_mat * (amp_factor ^ beta)


## 9. FORMAT FINAL RESULT TABLE

res <- as.data.table(as.table(score_mat))
setnames(res, c("protein_final", "protein_groups", "score"))
setorder(res, -score)  # Sort by decreasing score

res_final <- res[score >= score_cutoff & !is.na(score)]


## 10. EXPORT RESULTS AND PREVIEW

fwrite(res_final, file = "cosine_results.tsv", sep = "\t")
print(head(res_final))


```

11) Plot generation for individual protein pairs

-   Create a folder to store the output plots.

-   Define a utility function (get_pair_long) to:

    -   Extract the expression profile of a protein from df_spe_final.

    -   Extract the expression profile of a protein from df_proteo_merged.

    -   Combine both profiles into a single long-format table.

-   Iterate over all matched protein pairs from res_final:

    -   Generate a line plot comparing the two profiles across groups.

    -   Customize the plot aesthetics (title, axis labels, minimal theme).

    -   Save each plot as a .png file with a safe filename.

    -   Display a confirmation message once the file is saved.

```{r}

## 11. INDIVIDUAL PLOTS (res_final, sorted by decreasing score)

library(ggplot2)
library(dplyr)   # Load last to avoid conflicts
library(tidyr)
library(purrr)

# Output directory for plots
out_dir <- "pair_plots_sans_correction"
dir.create(out_dir, showWarnings = FALSE)

# Utility function: extract data for a given protein pair
get_pair_long <- function(prot_f, prot_g) {

  # Extract expression profile from df_spe_final
  v_final <- df_spe_final %>%
    dplyr::filter(protein == prot_f) %>%
    dplyr::select(dplyr::starts_with("Group")) %>%
    tidyr::pivot_longer(
      cols      = everything(),
      names_to  = "group",
      values_to = "value"
    ) %>%
    dplyr::mutate(source = prot_f)

  # Extract expression profile from df_proteo_merged
  v_groups <- df_proteo_merged %>%
    dplyr::select(group, dplyr::all_of(prot_g)) %>%
    dplyr::rename(value = !!rlang::sym(prot_g)) %>%
    dplyr::mutate(source = prot_g)

  # Merge both profiles into a single long-format table
  dplyr::bind_rows(v_final, v_groups) %>%
    dplyr::mutate(group = factor(tolower(group), levels = paste0("group", 1:5)))
}

# Generate plots for all protein pairs
purrr::pwalk(
  .l = list(res_final$protein_final, res_final$protein_groups),
  .f = function(prot_f, prot_g) {

    dat <- get_pair_long(prot_f, prot_g)

    p <- ggplot(dat, aes(group, value, colour = source, group = source)) +
      geom_line(size = 1) +
      geom_point(size = 2) +
      labs(
        title  = paste(prot_f, "vs", prot_g),
        x      = "Group",
        y      = "Value",
        colour = "Protein"
      ) +
      theme_minimal(base_size = 12)

    # Safe filename (replace special characters with underscores)
    safe <- function(x) gsub("[^A-Za-z0-9]", "_", x)
    file_name <- sprintf("%s/%s_vs_%s.png", out_dir, safe(prot_f), safe(prot_g))

    # Save the plot
    ggsave(file_name, plot = p, width = 10, height = 4, dpi = 300, bg = "white")
    message("✔  ", file_name)
  }
)

```

## 2) top 300 different

Runs the same similarity check on the intensity-driven protein list.

```{r}
#df_top_expressed_final <- bind_rows(result_list_2)   
```

```{r}
#suppressPackageStartupMessages(library(dplyr))   # Load dplyr last to avoid conflicts

## 1) Extract matrices ---------------------------------------------------------
#grp_cols  <- grep("^Group\\d+$", names(df_top_expressed_final), value = TRUE)
#mat_final_top <- as.matrix(df_top_expressed_final[, grp_cols])
#rownames(mat_final_top) <- df_top_expressed_final$protein

#group_names <- tolower(df_proteo_merged[[1]])           # "group1"…"group5"
#mat_groups  <- t(as.matrix(df_proteo_merged[, -1]))
#rownames(mat_groups) <- colnames(df_proteo_merged)[-1]
#colnames(mat_groups) <- group_names

## 2) Harmonize matrices -------------------------------------------------------
#col_order           <- tolower(grp_cols)       # "group1"…"group5"
#colnames(mat_final_top) <- col_order
#mat_groups          <- mat_groups[, col_order, drop = FALSE]
```

```{r}

## 0. PARAMETERS TO ADJUST

#min_sd         <- 0.50   # Minimum standard deviation to keep a protein
#score_cutoff   <- 0.00   # Minimum combined score threshold
#beta           <- 0      # Weight for the amplitude bonus
#use_differences <- FALSE # TRUE = use differences (slopes) instead of raw values
#center_only     <- FALSE # TRUE = center values without scaling to unit variance


## 1. LOAD REQUIRED PACKAGES

#suppressPackageStartupMessages(library(data.table))


## 2. ALIGN COLUMNS

# Keep only common group columns between the two matrices
#common_groups <- intersect(colnames(mat_final_top), colnames(mat_groups))
#mat_final_top_raw  <- as.matrix(mat_final_top[, common_groups])
#mat_groups_raw <- as.matrix(mat_groups[, common_groups])


## 3. FILTER BASED ON STANDARD DEVIATION

# Remove proteins with low variability
#keep_final  <- apply(mat_final_top_raw,  1, sd, na.rm = TRUE) > min_sd
#keep_groups <- apply(mat_groups_raw, 1, sd, na.rm = TRUE) > min_sd

#mat_final_top  <- mat_final_top_raw[keep_final , , drop = FALSE]
#mat_groups <- mat_groups_raw[keep_groups, , drop = FALSE]


## 4. OPTIONAL: COMPUTE DIFFERENCES (SLOPES)

#if (use_differences) {
  mat_final_top  <- t(apply(mat_final_top,  1, diff))  # n x (g-1)
  mat_groups <- t(apply(mat_groups, 1, diff))
}


## 5. NORMALIZATION: CENTERING OR SCALING

#if (center_only) {
 # center_rows <- function(m) t(apply(m, 1, function(v) v - mean(v, na.rm = TRUE)))
 # X <- center_rows(mat_final_top)
  #Y <- center_rows(mat_groups)
#} else {
  #scale_rows <- function(m) t(apply(m, 1, scale))
 # X <- scale_rows(mat_final_top)
 # Y <- scale_rows(mat_groups)
#}


## 6. SHAPE SIMILARITY (COSINE SIMILARITY)

#num   <- X %*% t(Y)
#denom <- outer(sqrt(rowSums(X^2)), sqrt(rowSums(Y^2)))
#sim_mat <- num / denom  # n_final × n_groups matrix


## 7. AMPLITUDE BONUS

#sd_final  <- apply(mat_final,  1, sd)
#sd_groups <- apply(mat_groups, 1, sd)

#amp_sim    <- function(s1, s2) pmin(s1, s2) / pmax(s1, s2)
#amp_factor <- outer(sd_final, sd_groups, amp_sim)  # Values between 0 and 1


## 8. COMBINED SCORE (SHAPE × AMPLITUDE^β)

#score_mat <- sim_mat * (amp_factor ^ beta)


## 9. FORMAT FINAL RESULT TABLE

#res <- as.data.table(as.table(score_mat))
#setnames(res, c("protein_final", "protein_groups", "score"))
#setorder(res, -score)  # Sort by decreasing score

#res_final <- res[score >= score_cutoff & !is.na(score)]


## 10. EXPORT RESULTS AND PREVIEW

#fwrite(res_final, file = "cosine_results_top.tsv", sep = "\t")
#print(head(res_final))
```

```{r}

## 11. INDIVIDUAL PLOTS (res_final, sorted by decreasing score)

#library(ggplot2)
#library(dplyr)   # Load last to avoid conflicts
#library(tidyr)
#library(purrr)

# Output directory for plots
#out_dir <- "pair_plots_sans_correction"
#dir.create(out_dir, showWarnings = FALSE)

# Utility function: extract data for a given protein pair
#get_pair_long <- function(prot_f, prot_g) {

  # Extract expression profile from df_top_expressed_final
  #v_final <- df_top_expressed_final %>%
   # dplyr::filter(protein == prot_f) %>%
   # dplyr::select(dplyr::starts_with("Group")) %>%
    #tidyr::pivot_longer(
    #  cols      = everything(),
     # names_to  = "group",
    #  values_to = "value"
  #  ) %>%
  #  dplyr::mutate(source = prot_f)

  # Extract expression profile from df_proteo_merged
#  v_groups <- df_proteo_merged %>%
 #   dplyr::select(group, dplyr::all_of(prot_g)) %>%
 #   dplyr::rename(value = !!rlang::sym(prot_g)) %>%
  #  dplyr::mutate(source = prot_g)

  # Merge both profiles into a single long-format table
  #dplyr::bind_rows(v_final, v_groups) %>%
   # dplyr::mutate(group = factor(tolower(group), levels = paste0("group", 1:5)))
#}

# Generate plots for all protein pairs
#purrr::pwalk(
 # .l = list(res_final$protein_final, res_final$protein_groups),
  #.f = function(prot_f, prot_g) {

   # dat <- get_pair_long(prot_f, prot_g)

    #p <- ggplot(dat, aes(group, value, colour = source, group = source)) +
     # geom_line(size = 1) +
      #geom_point(size = 2) +
      #labs(
       # title  = paste(prot_f, "vs", prot_g),
        #x      = "Group",
        #y      = "Value",
        #colour = "Protein"
      #) +
      #theme_minimal(base_size = 12)

    # Safe filename (replace special characters with underscores)
    #safe <- function(x) gsub("[^A-Za-z0-9]", "_", x)
    #file_name <- sprintf("%s/%s_vs_%s.png", out_dir, safe(prot_f), safe(prot_g))

    # Save the plot
   # ggsave(file_name, plot = p, width = 10, height = 4, dpi = 300, bg = "white")
    #message("✔  ", file_name)
#  }
#)
```
